\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2020-02-06 analysis/paper/paper.Rmd

eval_introduction <- FALSE
eval_theory <- FALSE
eval_implementation <- FALSE
eval_cs1 <- FALSE
eval_cs2 <- FALSE
eval_cs3 <- TRUE
eval_cs4 <- FALSE
```

@@

For the third case study, we will introduce regularization into the estimation of the hazard function. This can be performed using a combination of case-base sampling and regularized logistic regression. To demonstrate this capability, we examine the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) dataset [@knaus1995support]. The SUPPORT dataset tracks death for individuals who are considered seriously ill within a hospital. Imputation was conducted using the \pkg{mice} package in \proglang{R} with default settings [@mice]. Before imputation, there were a total of 1000 individuals and 35 variables. charges, Totcst, totmcst and sfdm2 were removed as they are response variables we are not interested in for the sake of this analysis. After imputation, we have 690 individuals. We lose individuals due to coliniarity between variables, preventing imputation. If a pair of covariates are deemed colinear to a certain degree, they are excluded from any imputations. Here, only edu  income  avtisst*  race  ph*  glucose*  adlp* were imputed. Of these 690 individuals, 66.09% experienced the event of interest. <!--The SUPPORT dataset will be used to demonstrate regularization within casebase, while comparing their absolute risk predictions for a new individual.--> A description of each variable can be found in Table \ref{tab:support1} and a breakdown of each categorical variable in Table \ref{tab:support2}. The original data was obtained from the Department of Biostatistics at Vanderbilt University. The imputed dataset is available as part of the \pkg{casebase} package.

@@

```{r support-poptime, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:support-poptime} Population-time plot for the SUPPORT data. Red points represent the case series. The gray space demonstrates the base series."}
data(support)
support_popTime <- popTime(support,
                           time = "d.time", event = "death",percentile_number = 0.31)
plot(support_popTime)
```

@@

```{r supportSetUp, echo = FALSE, eval = eval_cs3}
completeData=casebase::support
#remove hospital dead
completeData<-completeData[,-c(4)]
ratio=100  #universal ratio for eval_cs3
#keep one individual out, at random, for which we will develop an 
#absolute risk curve.
#sam=sample(1:nrow(completeData), 1)
#Set workingData to the remaining individuals in the data
#workingData=completeData[-c(sam),]
workingCompleteData=model.matrix(death~ .-d.time,data=completeData)[,-c(1)]#remove intercept
#Create u and xCox, which will be used when fitting Cox with glmnet.
newData=workingCompleteData
x=workingCompleteData
#x and y will be used to fit the casebase model under glmnet.
#x=as.matrix(sparse.model.matrix(death~ .-d.time+0,data=workingData))
y=data.matrix(completeData[,c(2,5)])
```

Our main objective is to compute the absolute risk of death for a given set of covariates, using regularization when fitting our model. First, we fit a smooth hazard to the data using a weibull distribution. As casebase makes use of the glmnet package, we interact with the fitsmoothhazard.fit function using a matrix interface, where y contains the time and event variables and x contains all other variables we would like to include in the model. For the SUPPORT dataset, all factor variables were converted into indicator variables. Here, we used lasso penalization by setting alpha to 1.

```{r cbHazardInteractions}, echo = TRUE, eval = eval_cs3}

cb.Model=casebase::fitSmoothHazard.fit(x,y,family="glmnet",time="d.time",event="death",
                                       formula_time = ~log(d.time),alpha=1,ratio=ratio)
```
Then, we take our model and use the `absoluteRisk` function to integrate and retrieve our cumulative incidence curves.

```{r CBabsolute, echo = FALSE, eval = eval_cs3}
#Estimating the absolute risk curve using the newData parameter.
casebaseCI=casebase::absoluteRisk(cb.Model,time = seq(1,max(completeData$d.time), 20),newdata = x , s="lambda.1se",method=c("numerical"))

cbCIAve=data.frame(casebaseCI[,1],rowMeans(casebaseCI[,-c(1)]))
```
We will compare this curve to a regularized version of cox regression, and a kaplan-meier survival curve. The cox regression absolute risk curve with lasso penalization required two extra steps to retrieve. Coxnet from the glmnet package has tools to fit a regularized cox model, but requires a survival object skeleton of the selected variables, so that the survival package tools can be used to retrive the absolute risk. We handle this by fitting a regularized cox model with the glmnet package, and a cox model from the survival package with the selected variables from the regularized fit. This second model serves as a skeleton that permits the use of survival package functions. The coefficients from the regularized model will replace the coefficients in our skeleton. the resulting model will have the correct coefficients when integrating, but will have an incorrect standard error. For the purposes of this case study, we are only interested in the absolute risk curve itself and not the standard error.

```{r coxHazAbsolute, echo = FALSE, eval = eval_cs3}
#Create u and xCox, which will be used when fitting Cox with glmnet.
u=survival::Surv(time = as.numeric(y[,2]), event = as.factor(y[,1]))
xCox=as.matrix(sparse.model.matrix(death~ .-d.time,data=completeData))
#hazard for cox using glmnet
coxglmFit=glmnet::cv.glmnet(x=xCox,y=u, family="cox",alpha=1)
#convergence demonstrated in plot
#plot(coxglmFit)
#taking the coefficient estimates for later use
nonzero_covariate_cox <- predict(coxglmFit, type = "nonzero", s = "lambda.1se")
nonzero_coef_cox <- coef(coxglmFit, s = "lambda.1se")
#creating a new dataset that only contains the covariates chosen through glmnet.
#cleanCoxData<- as.data.frame(cbind(as.numeric(workingData$d.time),as.factor(workingData$death), xCox[,nonzero_covariate_cox$X1]))
cleanCoxData<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCox[,nonzero_covariate_cox$X1]))
#newDataCox<-xCox[sam,nonzero_covariate_cox$X1]
#fitting a cox model using regular estimation, however we will not keep it.
#this is used more as an object place holder.
coxFit <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = cleanCoxData)
#The coefficients of this object will be replaced with the estimates from coxglmFit.
#Doing so makes it so that everything is invalid aside from the coefficients.
#In this case, all we need to estimate the absolute risk is the coefficients.
#Std. error would be incorrect here, if we were to draw error bars.
coxFit$coefficients<-nonzero_coef_cox@x
#Fitting absolute risk curve for cox+glmnet. -1 to remove intercept
newDataCox=newData[,nonzero_covariate_cox$X1-1]
abCoxFit<-survival::survfit(coxFit,newdata=as.data.frame(newDataCox),time = seq(0,max(completeData$d.time), 1),type="breslow")
abCF<-abCoxFit
abCF$surv<-rowMeans(abCF$surv)
```
We used the survival package to calculate cox regression's cumulative incidience without penalization and the kaplan-meier curve for incidience.

```{r unpenalizedCox, echo = FALSE, eval = eval_cs3}
Cox<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCox[,-c(1)]))
#coxFitup <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = Cox)
#abcoxFitup<-survival::survfit(coxFitup,newdata=Cox)
#abCFup<-abcoxFitup
#abCFup$surv<-rowMeans(abCFup$surv)
#cb.ModelNorm=casebase::fitSmoothHazard(V2~.,time="V1",event="V2",
       #                                ratio=rat,data=fullDataCox)
#summary(cb.ModelNorm)#need to figure out how to deal with NAs
#casebaseAbsoluteNorm=casebase::absoluteRisk(cb.ModelNorm,time = seq(1,max(completeData$d.time), 200),newdata = newData , s="lambda.1se",method=c("numerical"))
#cabn=data.frame(casebaseAbsoluteNorm[,1],rowMeans(casebaseAbsoluteNorm[,-c(1)]))
```

@@

Now that our four cumulative incidence estimations have been calculated, we compare them all in Figure \ref{fig:abSC}. Both coxnet and casebase decrease the absolute risk by the end of the study, in comparison to kaplan-meier.
```{r abSupportComparison, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:abSC} Compares regularized casebase in black, regularized cox in red and Kaplan meier in blue."}
abCFup<-abcoxFitup
abCFup$surv<-rowMeans(abCFup$surv)
# A plot to compare all three Absolute risk (Commulative Incidence)
plot(cbCIAve[,1],cbCIAve[,2],type='l',col="black",lwd=3,main="Cumulative Incidence comparison",xlab="Survival-Time",ylab="Cumulative Incidence",ylim=c(0,1),xlim=)
lines(abCF,col="red",fun="event",lwd=3,conf.int = FALSE)
lines(abKm ,col="Blue",fun="event",lwd=3,conf.int = FALSE)
lines(abCFup ,col="purple",fun="event",lwd=3,conf.int = FALSE)
legend("bottomright", 
       legend = c( "Penalized Weibull with casebase","Coxnet","Cox","Kaplan-Meier"), 
       col = c("black","red","Blue","purple"),
       lty = c(1, 1, 1), 
       bg = "gray90")
```

@@

#get covariates excluding d.time
estimatescb=setDT(as.data.frame(coef(cb.Model)[-c(1,2),1]), keep.rownames = TRUE)[]
estimatescb$Model="casebase"
colnames(estimatescb)<-c("Coefs","Estimate","Model")
estimatescox=setDT(as.data.frame(coef(coxglmFit)[-c(1),1]), keep.rownames = TRUE)[]
estimatescox$Model="Cox"
colnames(estimatescox)<-c("Coefs","Estimate","Model")
estimates=rbind(estimatescb, estimatescox)
estimates$Estimate[estimates$Estimate==0]=NA# make it so unchosen covariates appear empty
library(ggstance)# for vertical dodge
  ggplot(estimates, aes(x = Estimate, y = Coefs, color = Model)) +
        geom_segment(aes(x = 0, y = Coefs, xend = Estimate, yend = Coefs),size=0.5 ) +
        geom_point(position=ggstance::position_dodgev(height=0.2))

>>>>

eval_introduction <- FALSE
eval_theory <- TRUE
eval_implementation <- FALSE
eval_cs1 <- FALSE
eval_cs2 <- FALSE
eval_cs3 <- TRUE
eval_cs4 <- FALSE
```

@@

For the third case study, we will look at regularized estimation for the hazard function. This can be performed using a combination of case-base sampling and regularized logistic regression. To demonstrate this capability, we examine the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) dataset [@knaus1995support]. The SUPPORT dataset tracks death for individuals who are considered seriously ill within a hospital. Imputation was conducted using the \pkg{mice} package in \proglang{R} with default settings [@mice]. Before imputation, there were a total of 1000 individuals and 35 variables. After imputation, we have 690 individuals. We lose individuals due to coliniarity. If a pair of covariates are deemed colinear to a certain degree, they are excluded from any imputations. Here, only edu  income  avtisst*  race  ph*  glucose*  adlp* were imputed. Of these, 66.09% experienced the event of interest. <!--The SUPPORT dataset will be used to demonstrate regularization within casebase, while comparing their absolute risk predictions for a new individual.--> A description of each variable can be found in Table \ref{tab:support1} and a breakdown of each categorical variable in Table \ref{tab:support2}. The original data was obtained from the Department of Biostatistics at Vanderbilt University. The imputed dataset is available as part of the \pkg{casebase} package.

@@

```{r support-poptime, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:support-poptime} Population-time plot for the SUPPORT data."}
data(support)
support_popTime <- popTime(support,
                           time = "d.time", event = "death",percentile_number = 0.31)
plot(support_popTime)
```

@@

```{r supportSetUp, echo = FALSE, eval = eval_cs3}
completeData=casebase::support
#remove hospital dead
completeData<-completeData[,-c(4)]
rat=1  #universal ratio for eval_cs3
#keep one individual out, at random, for which we will develop an 
#absolute risk curve.
#sam=sample(1:nrow(completeData), 1)
#Set workingData to the remaining individuals in the data
#workingData=completeData[-c(sam),]
workingCompleteData=model.matrix(death~ .-d.time,data=completeData)[,-c(1)]#remove intercept
#Create u and xCox, which will be used when fitting Cox with glmnet.
newData=workingCompleteData
x=workingCompleteData
#x and y will be used to fit the casebase model under glmnet.
#x=as.matrix(sparse.model.matrix(death~ .-d.time+0,data=workingData))
y=data.matrix(completeData[,c(2,5)])
```

Our main objective is to compute the absolute risk of death for a given set of covariates, using regularization when fitting our model. First, we fit a smooth hazard to the data; for the sake of this example, we opted for a linear term for time. As casebase makes use of the glmnet package, we interact with the fitsmoothhazard.fit function using a matrix interface, where y contains the time and event variables and x contains all other variables we would like to include in the model. For the SUPPORT dataset, all factor variables were converted into dummy variables. Here, we used lasso penalization by setting alpha to 1.

```{r cbHazardInteractions}, echo = TRUE, eval = eval_cs3}
xInteractions=x*y[,2]
colnames(xInteractions) <- paste("d.time*", colnames(xInteractions), sep = "")
xFull=cbind(x,xInteractions)
cb.ModelInter=casebase::fitSmoothHazard.fit(xFull,y,family="glmnet",time="d.time",event="death",
                                       formula_time = ~d.time,alpha=1,ratio=rat)
```
Then, we take our model and use the absoluteRisk function to integrate and retrieve our absolute risk function.

```{r CBabsoluteInteracts, echo = TRUE, eval = eval_cs3}
#Estimating the absolute risk curve using the newData parameter.
casebaseAbsoluteInter=casebase::absoluteRisk(cb.ModelInter,time = seq(1,max(completeData$d.time), 400),newdata = xFull , s="lambda.1se",method=c("numerical"))

cabI=data.frame(casebaseAbsoluteInter[,1],rowMeans(casebaseAbsoluteInter[,-c(1)]))
```
We will compare this curve to a regularized version of cox regression, and a kaplan-meier survival curve. The cox regression absolute risk curve with lasso penalization required two extra steps to retrieve. Coxnet from the glmnet package has tools to fit a regularized cox model, but requires a survival object skeleton of the selected variables, so that the survival package tools can be used to retrive the absolute risk. We handle this by fitting a regularized cox model with the glmnet package, and a cox model from the survival package with the selected variables from the regularized fit. This second model serves as a skeleton that permits the use of survival package functions. The coefficients from the regularized model will replace the coefficients in our skeleton. the resulting model will have the correct coefficients when integrating, but will have an incorrect standard error. For the purposes of this case study, we are only interested in the absolute risk curve itself and not the standard error.

```{r coxHazAbsoluteInteracts, echo = TRUE, eval = eval_cs3}
#Create u and xCox, which will be used when fitting Cox with glmnet.
u=survival::Surv(time = as.numeric(y[,2]), event = as.factor(y[,1]))

xCoxI=as.matrix(cbind(data.frame("(Intercept)"=rep(1,each=690)),xFull))
colnames(xCoxI)[1]="(Intercept)"

#hazard for cox using glmnet
coxNetI=glmnet::cv.glmnet(x=xCoxI,y=u, family="cox",alpha=1)
#convergence demonstrated in plot
#plot(coxNet)
#taking the coefficient estimates for later use
nonzero_covariate_cox <- predict(coxNetI, type = "nonzero", s = "lambda.1se")
nonzero_coef_cox <- coef(coxNetI, s = "lambda.1se")
#creating a new dataset that only contains the covariates chosen through glmnet.
#cleanCoxData<- as.data.frame(cbind(as.numeric(workingData$d.time),as.factor(workingData$death), xCox[,nonzero_covariate_cox$X1]))
cleanCoxDataI<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCoxI[,nonzero_covariate_cox$X1]))
#newDataCox<-xCox[sam,nonzero_covariate_cox$X1]
#fitting a cox model using regular estimation, however we will not keep it.
#this is used more as an object place holder.
coxNetFitI <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = cleanCoxDataI)
#The coefficients of this object will be replaced with the estimates from coxNet.
#Doing so makes it so that everything is invalid aside from the coefficients.
#In this case, all we need to estimate the absolute risk is the coefficients.
#Std. error would be incorrect here, if we were to draw error bars.
coxNetFitI$coefficients<-nonzero_coef_cox@x
#Fitting absolute risk curve for cox+glmnet. -1 to remove intercept
#posCox=nonzero_covariate_cox$X1
#newDataCoxI=cleanCoxDataI[,posCox]

abcoxNetI<-survival::survfit(coxNetFitI,type="breslow",newdata=cleanCoxDataI)

abCNI<-abcoxNetI
abCNI$surv=rowMeans(abCNI$surv)
```

```{r unpenalizedCox, echo = TRUE, eval = eval_cs3}
fullDataCox<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCoxI[,-c(1)]))
coxFitI <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = fullDataCox)
abcoxFitI<-survival::survfit(coxFitI,newdata=fullDataCox)
abCFI<-abcoxFitI
abCFI$surv<-rowMeans(abCFI$surv)
#cb.ModelNorm=casebase::fitSmoothHazard(V2~.,time="V1",event="V2",
       #                                ratio=rat,data=fullDataCox)
#summary(cb.ModelNorm)#need to figure out how to deal with NAs
#casebaseAbsoluteNorm=casebase::absoluteRisk(cb.ModelNorm,time = seq(1,max(completeData$d.time), 200),newdata = newData , s="lambda.1se",method=c("numerical"))
#cabn=data.frame(casebaseAbsoluteNorm[,1],rowMeans(casebaseAbsoluteNorm[,-c(1)]))

```
We used the survival package to calculate the kaplan-meier absolute risk function.

@@

Now that our four cumulative incidence estimations have been calculated, we compare them all in Figure \ref{fig:abSC}. Both coxnet and casebase decrease the absolute risk by the end of the study, in comparison to kaplan-meier.
```{r abSupportComparison, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:abSC} Compares regularized casebase in black, regularized cox in red and Kaplan meier in blue."}
abCFup<-abcoxFitup
abCFup$surv<-rowMeans(abCFup$surv)
# A plot to compare all three Absolute risk (Commulative Incidence)
plot(cabI[,1],cabI[,2],type='l',col="black",lwd=3,main="Support- CaseBase vs. Cox+glmnet vs KM Absolute Risk Curves",xlab="Survival-Time",ylab="Cumulative Incidence",ylim=c(0,1),xlim=)
lines(abCFI,col="red",fun="event",lwd=3,conf.int = FALSE)
lines(abKm ,col="Blue",fun="event",lwd=3,conf.int = FALSE)
lines(abCNI,col="purple",fun="event",lwd=3,conf.int = FALSE)
legend("bottomright", 
       legend = c( "Casebase (Lasso+linear)","semi-parametric Coxnet","KM curve","cox"), 
       col = c("black","red","Blue","purple"),
       lty = c(1, 1, 1), 
       bg = "gray90")
```

@@

#get covariates excluding d.time
estimatescb=setDT(as.data.frame(coef(cb.Model)[-c(1,2),1]), keep.rownames = TRUE)[]
estimatescb$Model="casebase"
colnames(estimatescb)<-c("Coefs","Estimate","Model")
estimatescox=setDT(as.data.frame(coef(coxNet)[-c(1),1]), keep.rownames = TRUE)[]
estimatescox$Model="Cox"
colnames(estimatescox)<-c("Coefs","Estimate","Model")
estimates=rbind(estimatescb, estimatescox)
estimates$Estimate[estimates$Estimate==0]=NA# make it so unchosen covariates appear empty
library(ggstance)# for vertical dodge
  ggplot(estimates, aes(x = Estimate, y = Coefs, color = Model)) +
        geom_segment(aes(x = 0, y = Coefs, xend = Estimate, yend = Coefs),size=0.5 ) +
        geom_point(position=ggstance::position_dodgev(height=0.2))

```{r covariatesLolliInteractions,eval=eval_cs3}
#get covariates excluding d.time
library(data.table)
estimatescb=setDT(as.data.frame(coef(cb.ModelInter)[-c(1,2),1]), keep.rownames = TRUE)[]
estimatescb$Model="casebase"
colnames(estimatescb)<-c("Coefs","Estimate","Model")
estimatescox=setDT(as.data.frame(coef(coxNetI)[-c(1),1]), keep.rownames = TRUE)[]
estimatescox$Model="Cox"
colnames(estimatescox)<-c("Coefs","Estimate","Model")
estimatesI=rbind(estimatescb, estimatescox)
estimatesI$Estimate[estimatesI$Estimate==0]=NA# make it so unchosen covariates appear empty
library(ggstance)# for vertical dodge
  ggplot(estimatesI, aes(x = Coefs, y = Estimate , color = Model)) +
        geom_segment(aes(y = 0, x = Coefs, yend = Estimate, xend = Coefs),size=0.5, position = position_dodge(width = 0.5)) +
        geom_point(position = position_dodge(width = 0.5))
